### 2018-05-17 papers/mcswiggen/McSwiggen-Thesis.tex

\documentclass[a4paper,12pt]{article}
\usepackage{enumerate}

@@

%% Sets page size and margins
\usepackage[a4paper,top=3cm,bottom=2cm,left=1.5in,right=1in,marginparwidth=1.75cm]{geometry}

@@

\usepackage[justification=centering,format=plain,font=it]{caption}
\usepackage{minted}
\usepackage{setspace}

@@

@@

\title{Functionally Verified B+ Trees}
\author{Brian McSwiggen\\
        Advised by Professor Andrew Appel}

\begin{document}
\maketitle

@@

\doublespacing

\section{Abstract}

In this thesis, I make progress toward a fully verified functional B+\,tree program in Coq. A verified functional B+\,tree would serve as the basis for the future verification of an imperative B+\,tree program through proving equivalence, which would have applications in databases and other systems. The primary focus of the properties and abstractions in this thesis is to examine and describe the theory of the B+\,tree cursor, which is central to the operations of a B+\,tree both functionally and imperatively. The functional implementation presented here covers the essential cursor operations as well as B+\,tree lookups and insertions, but does not include a delete function. The progress made toward verification includes the development of a complete abstract specification, the structural correctness of cursor creation and positioning, and the relation of cursors and B+\,tree operations to an element-list abstract cursor representation.

@@

@@

B+\,trees are an extension of the more commonly known B-trees, which are N-ary trees often used to organize data in file systems. The property that makes B+\,trees unique is that the \textit{values}, that is the data actually being held in the tree, are kept exclusively at the leaves. The interior nodes hold only keys and pointers to their children.

One interesting consequence of this property is that B+\,trees in practice can be augmented with a linked list between the leaf nodes connecting all of the values. This makes finding a value stored in the tree efficient (due to the tree structure) while still keeping range queries and iteration over the values easy (due to the linked list). Additionally, because B+\,trees don’t have to store as much data in the interior nodes, a single node can fit more pointers, meaning that the fanout can be larger and the tree needs fewer levels, so data can be accessed more quickly \cite{elmasri_navathe_2011}. A well-designed B+\,tree will store enough values per leaf that a single leaf node just fits into a single storage block, allowing the maximum possible amount of data retrieval with the fewest possible I/O operations.

@@

Security-critical software, while it does not always carry the same risk to human life, does carry the risk of compromising crucial cryptographic tools, data privacy and integrity, and more. Much such software is incredibly important to modern technology, but is also very intricate, hard to implement correctly, and hard to keep secure. Formal verification provides a means of reasoning about these tricky tools. In 2015, the Verified Software Toolchain group completed a full verification of an OpenSSL HMAC implementation, meaning that as long as SHA-256 is in fact cryptographically secure (which remains unproven), the HMAC is as well \cite{190894}.

@@

Functional program verification such as the work done by Appel, Filliâtre, and Letouzey serves a practical purpose both because it can help find bugs in current functional libraries (as Filliâtre and Letouzey did with the OCaml Set module based on AVL trees \cite{filliatre:hal-00150913}) and because it can be extracted from Coq into live OCaml code and used in practice. However, many practical applications rely on imperative software. As an example in the domain of search trees, Xiwen Chen of York University provided a verified concurrent imperative binary search tree which he had proven linearizable, i.e. that any operation performs correctly as if it executed immediately \cite{chen_ruppert_breugel_2013}. Although it is more difficult to reason about imperative programs than about purely functional ones, doing imperative verifications is crucial for practical application of formal verification.

@@

\begin{figure}[h]
    \begin{center}
		\includegraphics[]{b}
        \caption{Diagram of VST components \protect\cite{vst_2017}}
    \end{center}
\end{figure}

Separation logic is crucial not just to a proof about a particular program, such as our B+\,trees implementation, but also to the entire verification stack that it is based on. Since the mid-2000s, a group at Princeton University led by Professor Andrew Appel has been developing the \textit{Verified Software Toolchain} \cite{vst_2017}, a stack of verified components that allow proofs to be written about source programs and apply to the compiled program as well. Each component of the toolchain, from the Verifiable C program logic to Xavier Leroy’s CompCert C compiler to the machine language operational semantics (and all of the interfaces between them), is designed for verification and proven correct \cite{10.1007/978-3-642-19718-5_1}.

Although the scope of this thesis does not extend into connecting the B+\,tree specification to an imperative C program that it represents, the functional specification produced could be useful for the verification of a C program. The strategy of functional specification used in this thesis is intended to work with the Verified Software Toolchain, so that a C program could be proven via the VST separation logic to be equivalent to this functional specification.

\subsection{DeepSpec}

The Verified Software Toolchain itself is part of a larger research group called DeepSpec. Since 2016, the DeepSpec group at Princeton, University of Pennsylvania, Yale, and MIT have been collaborating under an NSF Expedition in Computing grant to explore the science of deep specifications. This thesis is in part an attempt to extend that work by applying deep specification to a B+\,tree database program. Deep specifications, as described by the DeepSpec group, are program specifications with four special properties \cite{deepspec}:

@@

The strategy that this thesis targets for verifying imperative software is through a \textit{functional model}. A functional model describes the actual behavior of the operations, and in this case the model is an equivalent piece of software written in a functional programming language, but other functional models are possible. When reasoning about a programming language, for example, the functional model might be the small-step operational semantics of the language. For the purposes of this thesis, since functional programs are much easier to formally reason about, the functional model provides a bridge: we can prove the equivalence of the functional program and the imperative program, and then any proofs of correctness for the functional program are also valid proofs of the imperative one.

@@

The particular style of B+\,tree and set of operations which are targeted in this work are derived from a subset of the SQLite commands. SQLite uses B+\,trees to store information in relational databases. One particularity of SQLite is that it represents a cursor into the B+\,tree as a list of pointers to the split point at each level. This makes it easier to change the position of the cursor without needing to have a linked list at the leaves as mentioned above or having to retraverse the tree.

@@

%Maybe come back and clean this up later%
\begin{table}[h]
\centering
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
$P_1$ & $K_1$ & $P_2$ & $K_2$ & ... & $K_{n-1}$ & $P_n$\\
\hline
\end{tabular}
\caption{B+\,tree interior node with $n-1$ keys and $n$ subtrees}
\label{tab:interior-node}
\end{table}

@@

%Maybe come back and clean this up later%
\begin{table}[h]
\centering
\begin{tabular}{|l|l|l|l|}
\hline
$(K_1,V_1)$ & $(K_2,V_2)$ & ... & $(K_n,V_n)$\\
\hline
\end{tabular}
\caption{B+\,tree leaf node with $n$ keys and $n$ values}
\label{tab:leaf-node}
\end{table}

@@

%Find a way to position better?%
%Find a way to italicize captions?%
\begin{figure}[h]
    \centering
    \begin{tikzpicture}
    \tikzstyle{bplus}=[rectangle split, rectangle split horizontal,rectangle split ignore empty parts,draw]
    \tikzstyle{every node}=[bplus]
    \tikzstyle{level 1}=[sibling distance=40mm]
    \tikzstyle{level 2}=[sibling distance=15mm]
    \node {7} [->]
      child {node {3}
        child {node {1 \nodepart{two} 2}}
        child {node {3 \nodepart{two} 4 \nodepart{three} 6}}    
      } 
      child {node {9 \nodepart{two} 11}
        child {node {7 \nodepart{two} 8}}
        child {node {9 \nodepart{two} 10}}   
        child {node {11 \nodepart{two} 12}}    
      }
    ;\end{tikzpicture}
    \caption{A simple B+\,tree of height 3. Each interior node has 2 or 3 children.}
    \label{fig:demotree1}
\end{figure}

@@

\subsection{Lookup}

@@

\begin{figure}[h]
    \centering
    \begin{tikzpicture}
    \tikzstyle{bplus}=[rectangle split, rectangle split horizontal,rectangle split ignore empty parts,draw]
    \tikzstyle{every node}=[bplus]
    \tikzstyle{level 1}=[sibling distance=40mm]
    \tikzstyle{level 2}=[sibling distance=15mm]
    \node {7} [->]
      child {node {3}
        child {node {1 \nodepart{two} 2}}
        child {node {3 \nodepart{two} 4 \nodepart{three} 6}}    
      } 
      child[ultra thick,red] {node {9 \nodepart{two} 11}
        child[thin, black] {node {7 \nodepart{two} 8}}
        child[ultra thick, red] {node {9 \nodepart{two} 10}}   
        child[thin, black] {node {11 \nodepart{two} 12}}    
      }
    ;\end{tikzpicture}
    \caption{To find 10, we search for 10 at each level and recursively descend.}
    \label{fig:demotree2}
\end{figure}

@@

Insertion is simple if the leaf node has less than $b$ children: we simply find the correct place according to the list of keys, and add the value there. However, we must maintain that any given node has at most b children. Therefore, if a key is inserted which would go into a leaf that has b children already, then we have to split that node and add the middle key to the node above \cite{elmasri_navathe_2011}.

@@

%Could use some work%
\begin{figure}[h]
    \centering
    \begin{tikzpicture}
    \tikzstyle{bplus}=[rectangle split, rectangle split horizontal,rectangle split ignore empty parts,draw]
    \tikzstyle{every node}=[bplus]
    \tikzstyle{level 1}=[sibling distance=50mm]
    \tikzstyle{level 2}=[sibling distance=15mm]
    \node {7} [->]
      child[ultra thick,red] {node {3 \nodepart{two} 5}
        child[thin, black] {node {1 \nodepart{two} 2}}
        child[ultra thick, red] {node {3 \nodepart{two} 4}}
        child[ultra thick, red] {node {5 \nodepart{two} 6}}    
      } 
      child {node {9 \nodepart{two} 11}
        child {node {7 \nodepart{two} 8}}
        child {node {9 \nodepart{two} 10}}   
        child {node {11 \nodepart{two} 12}}    
      }
    ;\end{tikzpicture}
    \caption{To add a mapping for 5, we have to first find where 5 goes in the tree, then add the new key-value pair and split nodes as necessary.}
    \label{fig:demotree3}
\end{figure}

@@

\begin{figure}[h]
    \centering
    \begin{tikzpicture}
    \tikzstyle{bplus}=[rectangle split, rectangle split horizontal,rectangle split ignore empty parts,draw]
    \tikzstyle{every node}=[bplus]
    \tikzstyle{level 1}=[sibling distance=40mm]
    \tikzstyle{level 2}=[sibling distance=20mm]
    \node {7} [->]
      child {node {3}
        child {node {1 \nodepart{two} 2}}
        child {node {3 \nodepart{two} 4 \nodepart{three} 6}}    
      } 
      child[ultra thick,red] {node {11}
        child[ultra thick, red] {node {7 \nodepart{two} 8 \nodepart{three} 10}}
        child[thin, black] {node {11 \nodepart{two} 12}}    
      }
    ;\end{tikzpicture}
    \caption{When we delete 9, we may need to merge nodes together.}
    \label{fig:demotree4}
\end{figure}

@@

Our abstract specification of a B+\,tree is that of a lookup table. That is, it stores data indexed by keys, and allows that data to be arbitrarily fetched, changed, added to, or removed. However, we add one more layer to this particular table abstraction. B+\,trees rely on a cursor, which could be any way of pointing to a particular place in the table. We build this notion into our abstract specification, allowing us to prove correctness relative to this abstract notion of a cursor in addition to a table. Here, the specification is shown as written in Coq.

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
Module Type CURSOR_TABLE.
 Parameter V: Type.
 Definition key := Z.
 Parameter table: Type.
 Parameter cursor: Type.
 Parameter empty_t: table.
\end{minted}
\end{singlespace}
\end{figure}

We parametrize our abstract cursor-table with a number of basic types. Any cursor-table must store a set of values of type \texttt{V}. It stores these within a table of type \texttt{table} (\texttt{empty\_t} being the most basic such table, storing no values) and accesses them using a cursor of type \texttt{cursor}. Any implementation of a cursor-table must provide these basic parameters, but the exact types are up to the specific implementation---this makes the module type abstract.

@@

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
 Parameter make_cursor: key -> table -> cursor.
 Parameter get_table: cursor -> table.
 Parameter get_key: cursor -> option key.
 Parameter get: cursor -> option V.
 Parameter insert: cursor -> key -> V -> cursor.
 Parameter next: cursor -> cursor.
 Parameter prev: cursor -> cursor.
 Parameter first_cursor: table -> cursor.
 Parameter last_cursor: table -> cursor.
\end{minted}
\end{singlespace}
\end{figure}

Any cursor-table must also define a set of operations on the table. Again, the specific implementations are left abstract: they are constrained only by their type signatures and the required axioms about them discussed below.

The most central of these functions is the \texttt{make\_cursor} operation, which takes any key and a table and creates a cursor at the appropriate location for that key in the table. Such a cursor is then used as input to most other operations. The reverse of \texttt{make\_cursor} is \texttt{get\_table}, which translate from a cursor back into the table it points into. We also require two shortcuts for creating cursors at the extremes of a table, these are \texttt{first\_cursor} and \texttt{last\_cursor} and are important in the definition of later axioms.

Once we have set a cursor, a cursor-table should be able to use it to do any of the desired operations: \texttt{get} to retrieve the value stored at the cursor, \texttt{get\_key} to retrieve the key mapped to that value, and \texttt{insert} to update the tree with a new key-value pair. Deletion is not included as part of this specification, but could be axiomatized similarly to insertion.

The cursor should also be movable from its position: \texttt{next} should move it to the next key greater than the cursor’s current position, while \texttt{prev} should move it to the last key lesser than its current position.

\begin{figure}[h]
\begin{singlespace}
\begin{minted}{coq}
 Parameter abs_rel: table -> cursor -> Prop.
 Parameter key_rel: key -> cursor -> Prop.
 Parameter eq_cursor : cursor -> cursor -> Prop.
 Parameter cursor_correct: cursor -> Prop.
 Parameter table_correct: table -> Prop.
\end{minted}
\end{singlespace}
\end{figure}

>>>>

\documentclass[12pt]{article}
\usepackage{enumerate}

@@

%% Sets page size and margins
\usepackage[top=3cm,bottom=2cm,left=1.5in,right=1in,marginparwidth=1.75cm]{geometry}

@@

\usepackage[justification=centering,format=plain,font=it]{caption}
\usepackage{minted}
\usepackage{xpatch,letltxmacro}
\LetLtxMacro{\cminted}{\minted}
\let\endcminted\endminted
\xpretocmd{\cminted}{\RecustomVerbatimEnvironment{Verbatim}{BVerbatim}{}}{}{}
\usepackage{setspace}
\usepackage[title]{appendix}
\usepackage{amsthm}

@@

\newcommand{\nocontentsline}[3]{}
\newcommand{\tocless}[2]{\bgroup\let\addcontentsline=\nocontentsline#1{#2}\egroup}

@@

%\title{Functionally Verified B+ Trees}
%\author{Brian McSwiggen\\
%        Advised by Professor Andrew Appel}

\begin{document}
\doublespacing

\thispagestyle{empty}
\begin{sc}
\vspace*{0in}
\begin{center}
    \LARGE The Theory and Verification of\\B+\,Tree Cursor Relations
\end{center}
\vspace{.5in}
\begin{center}
    Brian McSwiggen\\
    Advisor: Professor Andrew Appel
\end{center}
\vspace{.5in}
\begin{center}
   Submitted in partial fulfillment\\
of the requirements for the degree of\\
Bachelor of Science in Engineering

\includegraphics[width=0.25\textwidth]{logo.jpg}

Department of Computer Science\\
Princeton University\\
May 2018
\end{center}
\end{sc}

\clearpage
\pagenumbering{roman}

\begin{center}
    \LARGE \textbf{Abstract}
\end{center}

\textit{In this thesis, I make progress toward a fully verified functional B+\,tree program in the Coq proof assistant. A verified functional B+\,tree would serve as the basis for the future verification of an imperative B+\,tree program through proving equivalence, which would have applications in databases and other systems. The primary focus of the properties and abstractions covered in this thesis is to examine and describe the theory of the B+\,tree cursor, which is central to the operations of a B+\,tree both functionally and imperatively. The functional implementation presented here covers the essential cursor operations as well as B+\,tree lookups and insertions, but does not include a delete function. The progress made toward verification includes the development of a complete abstract specification, the structural correctness of cursor creation and positioning, and the relation of cursors and B+\,tree operations to an element-list abstract cursor representation.}

@@

\begin{center}
    \LARGE \textbf{Acknowledgements}
\end{center}

\begin{singlespace}
\small
\setlength{\parskip}{0.7em}
As my time at Princeton comes rushing to an end, I would like to give a heartfelt thanks to the many, many people who helped me make it, who carried me through this wild place, and who sometimes had to drag me kicking and screaming. I couldn’t possibly fit all of you on this page, but I’m going to make the font small and try my best.

First, to Professor Appel, I wouldn’t have any of this thesis or even know what formal verification is without having you as my advisor and my teacher. Thank you, for believing in me, for meeting with me every week, for giving me help and feedback as often as I asked for it, and for showing me all the cool things these tools can do. Also to Professor Walker, my second reader, thanks for teaching me functional programming in the first place.

To Aurèle and Tosin, thank you both for taking the time to explain things to me and answer my questions. Thank you also for letting an undergrad contribute as much as he can, even though you probably could have done my whole thesis, better, in half the time. I hope what I’ve done is useful to you anyways.

Thank you to all the professors and teachers and mentors and tutors and TAs that I have had throughout all my time in school, even before Princeton: Dr. McCall, Mr. Plummer, Madame Hecker, Dr. Kairet, Mrs. Mentor, and Mrs. Lewis, to name just a few.

To my two crazy families on campus---Theatre Intime and Quad---you may have driven me crazy from stress but you also made it all worth it. I can’t imagine a better place to be stuck working until 2am. I’m so proud of what we accomplished, and I have so much faith in the new officers to do even better.

To the Q33, forever in my heart, forever in my mind---but \textit{never} in the TV room.

Kika, you made my senior year wonderful, and you’ve been the best companion I could have asked for on this wild ride. I have full confidence that your senior year will be even more wonderful, more fruitful, and produce a better thesis than this one.

Some other people deserve a special shout-out for making sure I didn’t literally fail. Katie, for dragging me out of bed with Nutella for that god-awful 9am freshman physics precept. Julie, for placing bets on how long I could go without missing a class. Zach, for getting breakfast with me whenever I was worried I might sleep through a morning meeting. Swebb and Taylor, for always picking up the phone, even at ungodly hours of the night (or morning). Lydia, who has heard me rant about more stupid problems than anyone. Jillian, for knowing me since middle school, and for somehow remaining my friend even after finding out that I’m basically still a child.

I of course have to thank all the people who are already PTL, and all my friends in the Class of 2017, for proving that it can in fact be done. I’m not sure I would believe it otherwise.

Finally, to my family. My parents, for supporting me and guiding me (and even proofreading my thesis!). My brothers, for being role models, mentors, and friends. My Oma and Opa, for making it all possible.

Thank you all. I love you all so much.
\setlength{\parskip}{1em}
\end{singlespace}

@@

\begin{singlespace}
\vspace*{1.5in}
\begin{center}
    This thesis represents my own work in accordance with University regulations.
\end{center}
\begin{flushright}
    \textit{Brian McSwiggen}

    May 7, 2018
\end{flushright}
\end{singlespace}

\begin{singlespace}
\setlength{\parskip}{-0.1em}
\tableofcontents
\setlength{\parskip}{1em}
\end{singlespace}

\clearpage
\pagenumbering{arabic}

@@

B+\,trees are an extension of the more commonly known B-trees, which are N-ary trees often used to organize data in file systems. The property that makes B+\,trees unique is that the \textit{values}, that is the data actually being stored in the tree, are kept exclusively at the leaves. The interior nodes hold only keys and pointers to their children.

Databases and file systems often use B+\,trees, especially for indexing, because they can efficiently lookup a value and do a range query. Because B+\,trees don’t have to store as much data in the interior nodes, a single node can fit more pointers, meaning that the fanout can be larger and the tree needs fewer levels, so data can be accessed more quickly \cite{elmasri_navathe_2011}. A well-designed B+\,tree will store enough values per leaf that a single leaf node just fits into a single storage block, allowing the maximum possible amount of data retrieval with the fewest possible I/O operations. Additionally, because all of the values are in the leaves and not buried in higher nodes of the tree, it is relatively easy to do a query for all keys within a certain range. Some implementations augment the tree with a linked list between leaf nodes to make range queries even easier.

@@

Security-critical software, while it does not always carry the same risk to human life, does carry the risk of compromising crucial cryptographic tools, data privacy and integrity, and more. A lot of security-critical software is incredibly important to modern technology, but is also very intricate, hard to implement correctly, and hard to keep secure. Formal verification provides a means of reasoning about these tricky tools. In 2015, the Verified Software Toolchain group completed a full verification of an OpenSSL HMAC implementation, meaning that as long as SHA-256 is in fact cryptographically secure (which remains unproven), the HMAC is as well \cite{190894}.

@@

Functional program verification such as the work done by Appel, Filliâtre, and Letouzey serves a practical purpose both because it can help find bugs in current functional libraries (as Filliâtre and Letouzey did with the OCaml Set module based on AVL trees \cite{filliatre:hal-00150913}) and because it can be extracted from Coq into live OCaml code to be used in practice. However, many practical applications rely on imperative software. As an example in the domain of search trees, Xiwen Chen of York University provided a verified concurrent imperative binary search tree which he had proven linearizable, i.e. that any operation performs correctly as if it executed immediately \cite{chen_ruppert_breugel_2013}. Although it is more difficult to reason about imperative programs than about purely functional ones, doing imperative verifications is crucial for practical application of formal verification.

@@

\begin{figure}[ht]
    \begin{center}
		\includegraphics[]{b}
        \caption{Diagram of VST components \protect\cite{vst_2017}}
    \end{center}
\end{figure}

Separation logic is crucial not just to a proof about a particular program, such as this B+\,tree implementation, but also to the entire verification stack that it is based on. Since the mid-2000s, a group at Princeton University led by Professor Andrew Appel has been developing the \textit{Verified Software Toolchain} \cite{vst_2017}, a stack of verified components that allow proofs to be written about source programs and apply to the compiled program as well. Each component of the toolchain, from the Verifiable C program logic to Xavier Leroy’s CompCert C compiler to the machine language operational semantics (and all of the interfaces between them), is designed for verification and proven correct \cite{10.1007/978-3-642-19718-5_1}.

Although the scope of this thesis does not extend into connecting the B+\,tree specification to an imperative C program that it represents, the functional specification produced could be useful for the verification of a C program. The strategy of functional specification used in this thesis is intended to work with the Verified Software Toolchain, so that a C program could be proved via the VST separation logic to be equivalent to this functional specification.

\subsection{DeepSpec}
\label{subsec:deepspec}

The Verified Software Toolchain itself is part of a larger research group called DeepSpec. Since 2016, the DeepSpec group at Princeton, the University of Pennsylvania, Yale, and MIT have been collaborating under an NSF Expedition in Computing grant to explore the science of deep specifications. This thesis is in part an attempt to extend that work by applying deep specification to a B+\,tree database program. Deep specifications, as described by the DeepSpec group, are program specifications with four special properties \cite{deepspec}:

@@

The strategy that this thesis targets for verifying imperative software is through a \textit{functional model}. A functional model describes the actual behavior of the operations. In this case the model is an equivalent piece of software written in a functional programming language, but other functional models are possible. When reasoning about a programming language, for example, the functional model might be the small-step operational semantics of the language. For the purposes of this thesis, since functional programs are much easier to formally reason about, the functional model provides a bridge: we can prove the equivalence of the functional program and the imperative program, and then any proofs of correctness for the functional program are also valid proofs of the imperative one.

@@

The particular style of B+\,tree and set of operations which are targeted in this work are derived from a subset of the SQLite commands. SQLite uses B+\,trees to store information in relational databases. One particularity of SQLite is that it represents a cursor into the B+\,tree as a list of pointers to the split point at each level. This makes it easier to change the position of the cursor without needing to retraverse the tree.

@@

%Maybe come back and clean this up later%
\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|l|l|l|l|}
\hline
$P_1$ & $K_1$ & $P_2$ & $K_2$ & ... & $K_{n-1}$ & $P_n$\\
\hline
\end{tabular}
\caption{B+\,tree interior node with $n-1$ keys and $n$ subtrees}
\label{tab:interior-node}
\end{table}

@@

%Maybe come back and clean this up later%
\begin{table}[ht]
\centering
\begin{tabular}{|l|l|l|l|}
\hline
$(K_1,V_1)$ & $(K_2,V_2)$ & ... & $(K_n,V_n)$\\
\hline
\end{tabular}
\caption{B+\,tree leaf node with $n$ keys and $n$ values}
\label{tab:leaf-node}
\end{table}

@@

\begin{figure}[ht]
    \centering
    \begin{tikzpicture}
    \tikzstyle{bplus}=[rectangle split, rectangle split horizontal,rectangle split ignore empty parts,draw]
    \tikzstyle{every node}=[bplus]
    \tikzstyle{level 1}=[sibling distance=40mm]
    \tikzstyle{level 2}=[sibling distance=15mm]
    \node {7} [->]
      child {node {3}
        child {node {1 \nodepart{two} 2}}
        child {node {3 \nodepart{two} 4 \nodepart{three} 6}}    
      } 
      child {node {9 \nodepart{two} 11}
        child {node {7 \nodepart{two} 8}}
        child {node {9 \nodepart{two} 10}}   
        child {node {11 \nodepart{two} 12}}    
      }
    ;\end{tikzpicture}
    \caption{A simple B+\,tree of height 3. Each interior node has 2 or 3 children.}
    \label{fig:demotree1}
\end{figure}

@@

\subsection{Lookup}
\label{subsec:lookup}

@@

\begin{figure}[ht]
    \centering
    \begin{tikzpicture}
    \tikzstyle{bplus}=[rectangle split, rectangle split horizontal,rectangle split ignore empty parts,draw]
    \tikzstyle{every node}=[bplus]
    \tikzstyle{level 1}=[sibling distance=40mm]
    \tikzstyle{level 2}=[sibling distance=15mm]
    \node {7} [->]
      child {node {3}
        child {node {1 \nodepart{two} 2}}
        child {node {3 \nodepart{two} 4 \nodepart{three} 6}}    
      } 
      child[ultra thick,red] {node {9 \nodepart{two} 11}
        child[thin, black] {node {7 \nodepart{two} 8}}
        child[ultra thick, red] {node {9 \nodepart{two} 10}}   
        child[thin, black] {node {11 \nodepart{two} 12}}    
      }
    ;\end{tikzpicture}
    \caption{To find 10, we search for 10 at each level and recursively descend.}
    \label{fig:demotree2}
\end{figure}

@@

Insertion is simple if the leaf node has less than $b$ children: we simply find the correct place according to the list of keys, and add the value there. However, we must maintain that any given node has at most $b$ children. Therefore, if a key is inserted which would go into a leaf that has $b$ children already, then we have to split that node and add the middle key to the node above \cite{elmasri_navathe_2011}.

@@

%Could use some work%
\begin{figure}[ht]
    \centering
    \begin{tikzpicture}
    \tikzstyle{bplus}=[rectangle split, rectangle split horizontal,rectangle split ignore empty parts,draw]
    \tikzstyle{every node}=[bplus]
    \tikzstyle{level 1}=[sibling distance=50mm]
    \tikzstyle{level 2}=[sibling distance=15mm]
    \node {7} [->]
      child[ultra thick,red] {node {3 \nodepart{two} 5}
        child[thin, black] {node {1 \nodepart{two} 2}}
        child[ultra thick, red] {node {3 \nodepart{two} 4}}
        child[ultra thick, red] {node {5 \nodepart{two} 6}}    
      } 
      child {node {9 \nodepart{two} 11}
        child {node {7 \nodepart{two} 8}}
        child {node {9 \nodepart{two} 10}}   
        child {node {11 \nodepart{two} 12}}    
      }
    ;\end{tikzpicture}
    \caption{To add a mapping for 5, we have to first find where 5 goes in the tree, then add the new key-value pair and split nodes as necessary.}
    \label{fig:demotree3}
\end{figure}

@@

\begin{figure}[ht]
    \centering
    \begin{tikzpicture}
    \tikzstyle{bplus}=[rectangle split, rectangle split horizontal,rectangle split ignore empty parts,draw]
    \tikzstyle{every node}=[bplus]
    \tikzstyle{level 1}=[sibling distance=40mm]
    \tikzstyle{level 2}=[sibling distance=20mm]
    \node {7} [->]
      child {node {3}
        child {node {1 \nodepart{two} 2}}
        child {node {3 \nodepart{two} 4 \nodepart{three} 6}}    
      } 
      child[ultra thick,red] {node {11}
        child[ultra thick, red] {node {7 \nodepart{two} 8 \nodepart{three} 10}}
        child[thin, black] {node {11 \nodepart{two} 12}}    
      }
    ;\end{tikzpicture}
    \caption{When we delete 9, we may need to merge nodes together.}
    \label{fig:demotree4}
\end{figure}

@@

Our abstract specification of a B+\,tree is that of a lookup table. That is, it stores data indexed by keys, and allows that data to be arbitrarily fetched, changed, added to, or removed. However, we add one more layer to this particular table abstraction. B+\,trees rely on a cursor, which could be any way of pointing to a particular place in the table. We build this notion into our abstract specification, allowing us to prove correctness relative to this abstract notion of a cursor in addition to a table.

\textbf{Figure~\ref{fig:params}} and \textbf{Figure~\ref{fig:axioms}} show the specification as written in Coq. The \texttt{Module Type} describes the specification. An instance of \texttt{CURSOR\_TABLE} must provide a definition for every \texttt{Parameter} and a proof for every \texttt{Axiom}. The implementation developed in this thesis is described in \textbf{Section~\ref{sec:typedef}} and \textbf{Section~\ref{sec:func}}, and its relation to \texttt{CURSOR\_TABLE} is given in \textbf{Section~\ref{sec:proofs}}.

\begin{figure}[hbp]
\begin{singlespace}
\begin{minted}{coq}
Module Type CURSOR_TABLE.
 Parameter V: Type.
 Definition key := Z.
 Parameter table: Type.
 Parameter cursor: Type.
 Parameter empty_t: table.

 (* Functions of the implementation *)
 Parameter make_cursor: key -> table -> cursor.
 Parameter get_table: cursor -> table.
 Parameter get_key: cursor -> option key.
 Parameter get: cursor -> option V.
 Parameter insert: cursor -> key -> V -> cursor.
 Parameter next: cursor -> cursor.
 Parameter prev: cursor -> cursor.
 Parameter first_cursor: table -> cursor.
 Parameter last_cursor: table -> cursor.

 (* Predicates of the implementation *)
 Parameter abs_rel: table -> cursor -> Prop.
 Parameter key_rel: key -> cursor -> Prop.
 Parameter eq_cursor : cursor -> cursor -> Prop.
 Parameter cursor_correct: cursor -> Prop.
 Parameter table_correct: table -> Prop.
\end{minted}
\caption{The parameters in the first half of the Coq module type.}
\label{fig:params}
\end{singlespace}
\end{figure}

A cursor table is a relation that uses positionable cursors to access and update stored values of type \texttt{V}. As such, any cursor table implementation must have both a \texttt{table} type to represent an entire relation and a \texttt{cursor} type to act as a pointer into a relation.

\begin{figure}[p]
\begin{singlespace}
\begin{minted}{coq}
 (* Theorems about the predicates *)
 Axiom make_cursor_rel: forall t k,
   abs_rel t (make_cursor k t).
 Axiom get_table_rel: forall t c,
   abs_rel t c <-> get_table c = t.
 Axiom first_rel: forall t,
   abs_rel t (first_cursor t).
 Axiom last_rel: forall t,
   abs_rel t (last_cursor t).
 Axiom next_rel: forall t c,
   abs_rel t c -> abs_rel t (next c).
 Axiom prev_rel: forall t c,
   abs_rel t c -> abs_rel t (prev c).
 Axiom correct_rel: forall t c,
   abs_rel t c -> (cursor_correct c <-> table_correct t).

 (* Correctness preservation *)
 Axiom insert_correct: forall k v c,
   cursor_correct c -> key_rel k c -> cursor_correct (insert c k v).

 (* Get/insert correctness *)
 Axiom glast: forall t,
   get (last_cursor t) = None.
 Axiom gis: forall k v c,
   cursor_correct c -> key_rel k c ->
   get (make_cursor k (get_table (insert c k v))) = Some v.
 Axiom gio: forall j k v c,
   cursor_correct c -> key_rel k c -> ~ key_rel j c ->
   get (make_cursor j (get_table (insert c k v))) =
   get (make_cursor j (get_table c)).

 (* Cursor movement correctness *)
 Axiom next_prev: forall c t,
   cursor_correct c -> abs_rel t c -> ~ (c = last_cursor t) ->
   eq_cursor c (prev (next c)).
 Axiom prev_next: forall c t,
   cursor_correct c -> abs_rel t c -> ~ (c = first_cursor t) ->
   eq_cursor c (next (prev c)).
 Axiom cursor_order: forall c k1 k2,
   cursor_correct c -> get_key c = Some k1 ->
   get_key (next c) = Some k2 -> lt_key k1 k2 = true.
End CURSOR_TABLE.
\end{minted}
\caption{The axioms in the second half of the Coq module type.}
\label{fig:axioms}
\end{singlespace}
\end{figure}

@@

As suggested by the name of the module type, cursors are central to the operation of any cursor table. Appropriately, \texttt{make\_cursor} is the most central operation to the cursor table. All the other operations listed in \textbf{Figure~\ref{fig:params}} rely on a cursor produced by either \texttt{make\_cursor} or the related \texttt{first\_cursor} and \texttt{last\_cursor}. The functions \texttt{first\_cursor} and \texttt{last\_cursor} are needed to creating a cursor pointing at the very beginning or very end of a list, and these are primarily needed to define later \texttt{Axiom}s about the operation of a cursor table. The inverse of \texttt{make\_cursor} is \texttt{get\_table}: it extracts from a cursor the table that it points into.

Once we have set a cursor, a cursor-table should be able to use it to do any of the desired operations: \texttt{get} to retrieve the value stored at the cursor, \texttt{get\_key} to retrieve the key mapped to that value, and \texttt{insert} to update the tree with a new key-value pair. Deletion is not included as part of this specification, but could be axiomatized similarly to insertion. The cursor should also be movable from its position: \texttt{next} should move it to the next key greater than the cursor’s current position, while \texttt{prev} should move it to the last key lesser than its current position.

Beyond the definition of the implementation itself, a series of predicates about cursors and tables must also be defined. The first, \texttt{abs\_rel}, defines what it means for a particular cursor to be a reference into a particular table, i.e. for the table and cursor to represent the same abstract relation. Of course, this could be as simple as the predicate that \texttt{get\_table} on the cursor returns the table, but we leave room for the particular implementation to provide a more complex cursor-table correspondence that makes later proofs easier (as is the case for this implementation, as described in \textbf{Section~\ref{sec:func}}). The second, \texttt{key\_rel}, defines what it means for a cursor to be positioned for a particular key $k$, i.e. that the key to the right of the cursor is greater than $k$ and the key to the left is at most $k$. The third, \texttt{eq\_cursor}, is a predicate for two cursors being abstractly equivalent even if they are not structurally equal.

The fourth and fifth predicates describe what it means for a cursor or table to be correct: since we don’t require every instance of the \texttt{cursor} and \texttt{table} types to be correct, it is possible to have a cursor or table that does not actually correspond to a valid instance of a \texttt{CURSOR\_TABLE}. For example, with B+\,trees it is possible to construct an unbalanced or unsorted tree, in which case the operations are not guaranteed to be correct. In this implementation, it would have been possible to make a more complicated B+\,tree type than the one described in \textbf{Section~\ref{sec:typedef}} and enforce the correctness properties on every tree instance. However, this would have made working with that type more difficult.
